---
title: Compound specific isotope analsyis of human and faunal samples from Sambaquis
  sites
author: "OEC"
date: "26/04/2023"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
Load packages

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
library(tidyr)
library(stringr)
library(Hmisc)
library(ggpubr)
library(googlesheets4)
library (devtools)

# Load colour palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

Create a theme to apply to all plots
```{r}
theme_Pub <- function(base_size=14, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "right",
               legend.direction = "vertical",
               legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="bold"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}
```

Load data from googlesheet https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512

```{r}
gs4_deauth()
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")
humans <-filter(data, group == "Human")
fauna <-filter(data, group == "Fauna")
Piacaguera <-filter(humans, Site == "Piaçaguera")
control <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="d13C_control groups")

```

Create a colour pallette and order to apply to all plots
```{r}
cols <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "terrestrial_fauna" = "#CC79A7", "marine_fauna" = "#0072B2")

cols1 <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "Freshwater" = "#999999", "Marine" = "#0072B2", "C3 Fauna" = "#F0E442", "C3 Humans" = "#D55E00", "C4_humans" = "#CC79A7") 
           
order <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "terrestrial_fauna", "marine_fauna")

order1 <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "Freshwater", "Marine", "C3 Fauna", "C3 Humans", "C4_humans") 

```


Conduct a PCA on the Larsen raw amino acids

```{r}

#Load Larsen on d13C values for reference 
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset primary producters
Larsen_pp <- filter(Larsen, PP == "Y")
#Select groups and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
Larsen_pp
```

Next conduct and visualize the PCA

```{r}
#Select just amino acids for PCA
all_AA <- Larsen_pp %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Plot PC1 vs PC2
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  theme_Pub () +
  coord_fixed (ratio = 1)
p

```

Calculate the AAn values which equals each values variance from the mean and repeat the analysis.

First organise the data and normalise to row mean. 

```{r}

Larsen_pp <- filter(Larsen, PP == "Y")
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
Larsen_pp <- Larsen_pp %>% rowwise() %>% mutate(row_mean = mean(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% rowwise() %>% 
  mutate(across(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C), ~ .x - row_mean))
Larsen_pp_n
```

Then perform the PCA and visualise the results. This creates Figure 1 in Larsen et al. 

```{r}
#extract the amino acid values
all_AA <- Larsen_pp_n %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp_n$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
  #scale_colour_manual(cbPalette)+
xlim(5,-5)+
 ylim(5,-5)+
  #theme_Pub ()+
  coord_fixed (ratio = 1)
p
```
Repeat with just key source aminos acids


```{r}

Larsen_pp <- filter(Larsen, PP == "Y")
Larsen_pp <- Larsen_pp %>% dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
Larsen_pp <- Larsen_pp %>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% rowwise() %>% 
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
Larsen_pp_n
#Conduct the PCA
all_AA <- Larsen_pp_n %>%  dplyr::select(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp_n$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  theme_Pub () +
  coord_fixed (ratio = 1)
p
```


Obtain the 13CAAn PCA values from the TRADITION dataset


Load TRADITION data from googlesheet https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512

```{r}

gs4_deauth()
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")
humans <-filter(data, group == "Human")
fauna <-filter(data, group == "Fauna")
Piacaguera <-filter(humans, Site == "Piaçaguera")

#Create a colour pallette and order to apply to all plots
cols <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "terrestrial_fauna" = "#CC79A7", "marine_fauna" = "#0072B2")

order <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "terrestrial_fauna", "marine_fauna")

```

```{r}

data <- data %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
data <- na.omit(data) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
data <- data%>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
data_n <-data %>% rowwise() %>% 
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
data_n

```

Create a grouping variable and bind dataframes

```{r}
Tradition_AA <- bind_rows (data_n, Larsen_pp_n)
Tradition_AA
```
Conduct PCA on combined dataset

```{r}

all_AA <- Tradition_AA%>%  dplyr::select(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Tradition_AA$Group, PCA$x)
PCAvalues
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot() +
  geom_point(data=PCAvalues, aes(x = PC1, y = PC2, fill= group), pch=21, colour='black', size=2) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  theme_Pub ()+
  coord_fixed (ratio = 1)
p
```



#Conduct LDA analysis


Prepare Larsen data for LDA analysis
```{r}
#Select just amino acids for PCA
library(caret)
library(tidyverse)
#Create a normalised training set from Larsen's primary producters
gs4_deauth()
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset non primary producers
Larsen_pp <- filter(Larsen, PP == "Y")
#Select group and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_pp <- na.omit(Larsen_pp)
Larsen_pp$Group  <-as.factor (Larsen_pp$Group)
#omit Fungi
Larsen_pp <-Larsen_pp %>% filter(Group!="Fungi")
#Larsen_pp <-Larsen_pp %>% filter(Group!="Seagrasses")
```


Normalise the amino acids in Larsen reference dataset 
```{r}
#Create a new coloumn containing the row mean of the key amino acid for each row. .
Larsen_pp <- Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(across(c( ile_d13C, leu_d13C,  lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
Larsen_pp_n <- Larsen_pp_n %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)#drop means 
print (Larsen_pp_n)

```

Partition the data and create train and test sets 

```{r}
ind <- sample(2, nrow(Larsen_pp_n),
              replace = TRUE,
              prob = c(0.8, 0.2))
train.data <- Larsen_pp_n[ind==1,]
test.data <- Larsen_pp_n[ind==2,]
train.data
```

Create LDA model
```{r}
library(MASS)
# Fit LDA model to the training data
lda_model <- lda(Group ~ ., data = train.data)
```

```{r}
# Calculate the percentage of variance explained by each linear discriminant
percent_variance_explained <- lda_model$svd^2 / sum(lda_model$svd^2) * 100
ld1_label <- paste0("Linear Discriminant 1 (", round(percent_variance_explained[1], 2), "%)")
ld2_label <- paste0("Linear Discriminant 2 (", round(percent_variance_explained[2], 2), "%)")
```

Evaluate the LDA  performance
```{r}
# Predict on the test data using the LDA model
lda_predictions <- predict(lda_model, test.data)

#lda_predictions$class
confusion_matrix <- confusionMatrix(lda_predictions$class, test.data$Group)
print(confusion_matrix)
```

Call arcaheological data and normalise


```{r}
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")|> as.data.frame()
data <- data %>%  dplyr::select(Group, group, category, Site, Sample_ID, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
data <- data %>% dplyr::rowwise() %>% dplyr::mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))

#Subtract the row mean from each of the row values to normalize
data <- data %>% dplyr::rowwise() %>% dplyr::mutate(across(c( ile_d13C, leu_d13C,  lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
data_n <- data %>%  dplyr::select(Group, group, category, Sample_ID,  Site, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)#drop means 
data_n$Group <-as.factor (data_n$Group)
print (data_n)
```

```{r}
# Predict the classes for the unknown samples using the LDA model
lda_Sambaqui_predictions <- predict(lda_model, newdata = data_n)
#Print the predicted classes for the unknown samples
predicted_Sambaqui_classes <- lda_Sambaqui_predictions$class
lda_Sambaqui_predictions
```
Create dataframes for plotting 

```{r}
Refs <- data.frame(LD1 = predict(lda_model, newdata = train.data)$x[, 1], LD2 = predict(lda_model, data = train.data)$x[, 2], Type = "Training", group = as.factor(train.data$Group))

Samples <- data.frame(Sample_ID= data_n$Sample_ID, LD1 = lda_Sambaqui_predictions$x[, 1], LD2 = lda_Sambaqui_predictions$x[, 2], Type = "Samples", group = data_n$Group, Site =data_n$Site, Group1 = data_n$group, category= data_n$category)

Fauna_LDA <- Samples %>% dplyr::filter (Group1 == "Fauna")
Human_LDA <- Samples %>%filter(Group1 == "Human")
Human_LDA

#Combined <- rbind (Refs, Samples)
#print (Combined)
```



#Find means for automatic labelling
``` {r}
#Find means for automatic labelling
means <- Refs %>% group_by(group) %>%
  summarise(dplyr::across(c(LD1, LD2), mean))
means
#means <- means %>%mutate(cats=c("Porcine", "Ruminant", "Dairy", "Freshwater", "Marine"))#Sh
```


Plot the LDA
```{r}
# Plot the combined data
p <- ggplot() +
 #geom_point(data=Refs, aes(x =LD1, y =LD2, fill = group), pch=22, colour = "black", size = 2) +
  stat_ellipse(geom= "polygon", data=Refs, alpha=0.1,
               aes (x = LD1, y = LD2, group = group),
              level = 0.68)+
  geom_point(data=Human_LDA, aes(x =LD1, y =LD2, fill = Site), pch=23, colour = "black", size = 3) +
  geom_point(data=Fauna_LDA, aes(x =LD1, y =LD2, fill = category), pch=21, colour = "black", size = 3) +
   scale_fill_manual(values=c(cols), breaks=c(order))+
  geom_text(data=means, aes(y=LD2, x=LD1,label=group), hjust=0, size=4, color="black")+
  labs(title = "", x = ld1_label, y = ld2_label, fill = "Group", shape = "Type") +
  #xlim(-3,3)+
 #ylim(-3,3)+
  theme_Pub()
  #coord_fixed(ratio = 1)
p
```


Plot LD2 against sulfur isotopes for Piacaguera
```{r}
Piacaguera_LDA  <- Samples %>%filter(Site == "Piaçaguera") %>%filter(category == "human")
Piacaguera_LDA
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")|> as.data.frame()
Piacaguera_d34S <- data %>%filter(Site == "Piaçaguera") %>%filter(category == "human")  %>% dplyr::select (Sample_ID, d34S, d13C)
Piacaguera_d34S
Piacguera <- merge(Piacaguera_d34S, Piacaguera_LDA, by = c("Sample_ID"))
Piacguera
```

Plot d34S against LDA2 
```{r}
# Plot the combined data
p <- ggplot() +
  geom_point(data=Piacguera, aes(x =d34S, y = LD2, fill = Site), pch=23, colour = "black", size = 3) +
   scale_fill_manual(values=c(cols), breaks=c(order))+
  #geom_text(data=means, aes(y=LD2, x=LD1,label=group), hjust=0, size=4, color="black")+
  labs(y=ld2_label, x=expression(delta^{34}*S[collagen]*"(\u2030)"))+
  #xlim(-3,3)+
 #ylim(-3,3)+
  theme_Pub()
  #coord_fixed(ratio = 1)
p
```
```








