---
title: Compound specific isotope analsyis of human and faunal samples from Sambaquis
  sites
author: "OEC"
date: "26/04/2023"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
Load packages

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
library(tidyr)
library(stringr)
library(Hmisc)
library(ggpubr)
library(googlesheets4)
library (devtools)

# Load colour palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

Create a theme to apply to all plots
```{r}
theme_Pub <- function(base_size=14, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "right",
               legend.direction = "vertical",
               legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="bold"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}
```

Load data from googlesheet https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512

```{r}
gs4_deauth()
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")
humans <-filter(data, group == "Human")
fauna <-filter(data, group == "Fauna")
Piacaguera <-filter(humans, Site == "Piaçaguera")
control <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="d13C_control groups")

```

Create a colour pallette and order to apply to all plots
```{r}
cols <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "terrestrial_fauna" = "#CC79A7", "marine_fauna" = "#0072B2")

cols1 <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "Freshwater" = "#999999", "Marine" = "#0072B2", "C3 Fauna" = "#F0E442", "C3 Humans" = "#D55E00", "C4_humans" = "#CC79A7") 
           
order <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "terrestrial_fauna", "marine_fauna")

order1 <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "Freshwater", "Marine", "C3 Fauna", "C3 Humans", "C4_humans") 

```


Conduct a PCA on the Larsen raw amino acids

```{r}

#Load Larsen on d13C values for reference 
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset primary producters
Larsen_pp <- filter(Larsen, PP == "Y")
#Select groups and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
Larsen_pp
```

Next conduct and visualize the PCA

```{r}
#Select just amino acids for PCA
all_AA <- Larsen_pp %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Plot PC1 vs PC2
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  theme_Pub () +
  coord_fixed (ratio = 1)
p

```

Calculate the AAn values which equals each values variance from the mean and repeat the analysis.

First organise the data and normalise to row mean. 

```{r}

Larsen_pp <- filter(Larsen, PP == "Y")
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
Larsen_pp <- Larsen_pp %>% rowwise() %>% mutate(row_mean = mean(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% rowwise() %>% 
  mutate(across(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C), ~ .x - row_mean))
Larsen_pp_n
```

Then perform the PCA and visualise the results. This creates Figure 1 in Larsen et al. 

```{r}
#extract the amino acid values
all_AA <- Larsen_pp_n %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp_n$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
  #scale_colour_manual(cbPalette)+
xlim(5,-5)+
 ylim(5,-5)+
  #theme_Pub ()+
  coord_fixed (ratio = 1)
p
```
Repeat with just key source aminos acids


```{r}

Larsen_pp <- filter(Larsen, PP == "Y")
Larsen_pp <- Larsen_pp %>% dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
Larsen_pp <- Larsen_pp %>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% rowwise() %>% 
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
Larsen_pp_n
#Conduct the PCA
all_AA <- Larsen_pp_n %>%  dplyr::select(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp_n$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  theme_Pub () +
  coord_fixed (ratio = 1)
p
```


Obtain the 13CAAn PCA values from the TRADITION dataset


Load TRADITION data from googlesheet https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512

```{r}

gs4_deauth()
data <- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Sambaquis_CSIA_AA_clean")
humans <-filter(data, group == "Human")
fauna <-filter(data, group == "Fauna")
Piacaguera <-filter(humans, Site == "Piaçaguera")

#Create a colour pallette and order to apply to all plots
cols <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "terrestrial_fauna" = "#CC79A7", "marine_fauna" = "#0072B2")

order <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "terrestrial_fauna", "marine_fauna")

```

```{r}

data <- data %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
data <- na.omit(data) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
data <- data%>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
data_n <-data %>% rowwise() %>% 
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
data_n

```

Create a grouping variable and bind dataframes

```{r}
Tradition_AA <- bind_rows (data_n, Larsen_pp_n)
Tradition_AA
```
Conduct PCA on combined dataset

```{r}

all_AA <- Tradition_AA%>%  dplyr::select(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Tradition_AA$Group, PCA$x)
PCAvalues
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot() +
  geom_point(data=PCAvalues, aes(x = PC1, y = PC2, color = group, size=1)) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") + 
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  #theme_Pub ()+
  coord_fixed (ratio = 1)
p
```



#Conduct LDA analysis

---
title: Compound specific isotope analsyis of human and faunal samples from Denamrk
  sites
author: "OEC"
date: "26/04/2023"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
Load packages

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
library(tidyr)
library(stringr)
library(Hmisc)
library(ggpubr)
library(googlesheets4)

# Load colour palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```



#1 Load data from googlesheet 

```{r}
# Read data from INTERNET
library('googlesheets4')
 #gs4_deauth() # if needed
 gs4_auth(email='oliver.craig@york.ac.uk') # only if needed
data <-read_sheet("https://docs.google.com/spreadsheets/d/1TloKBguOGJ6akvWmUg9JCZ4Vife_RNkY_0NOd-EQUEg/edit#gid=812854588", sheet="Data") |> as.data.frame()
#subset data
library(dplyr)
fauna <- data %>% dplyr::filter (group == "Fauna")
humans <- data %>% dplyr::filter (group == "Human")
all_data <- data %>% dplyr::filter (Country == "Denmark")

```
```{r}

#cols <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "terrestrial_fauna" = "#CC79A7", "marine_fauna" = "#0072B2")

#cols1 <- c("Jabuticabeira II" = "#56B4E9", "Moraes" = "#009E73", "Piaçaguera" = "#E69F00", "Freshwater" = "#999999", "Marine" = "#0072B2", "C3 Fauna" = "#F0E442", "C3 Humans" = "#D55E00", "C4_humans" = "#CC79A7") 
           
#order <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "terrestrial_fauna", "marine_fauna")

#order1 <- c("Jabuticabeira II", "Moraes", "Piaçaguera", "Freshwater", "Marine", "C3 Fauna", "C3 Humans", "C4_humans") 

```



#2 Plot the bulk isotope data by species and site. 

```{r}
#Creates a template for collagen15N against collagen13C
p <-ggplot()+
  labs(y=expression(delta^{15}*N[collagen]*"(\u2030)"), x=expression(delta^{13}*C[collagen]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-25,-8))+
  scale_y_continuous(position = "left",limits=c(5,20))+
  geom_point(data=humans, aes(y=d15N, x=d13C, fill=Culture), colour="black", pch=23, size=3) +
  geom_point(data=fauna, aes(y=d15N, x=d13C, fill=category), colour="black", pch=21, size=3) +
  #scale_fill_manual(values=c(cols), breaks=c(order))+
  theme(legend.title=element_text(size=10))+
  guides(fill=guide_legend(title="Samples"))+
  theme_bw()+
  coord_fixed(ratio = 1)
p



```

#3. Quality control check the CSIA data by plotting Pro against Hyp va Pro for each isotope. 

```{r}
#QC3 Plot 15NPro against 15NHypro 
p <- ggplot(data=data, aes(y=hyp_d15N, x=pro_d15N, fill=Batch))+ 
  labs(x=expression(delta^{15}*N[Pro]*"(\u2030)"), y=expression(delta^{15}*N[Hyp]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(5,25))+
  scale_y_continuous(position = "left",limits=c(5,25))+
  geom_point(colour="black", pch=21, size=3.5)+
  #geom_smooth(method=lm, se=FALSE)+
  geom_abline(intercept=c(0), slope= 1, linetype="solid")+
  theme_bw()+
  coord_fixed(ratio = 1)
p

#QC4 Plot 13CPro against 13CHypr
p <- ggplot(data=data, aes(y=hyp_d13C, x=pro_d13C, fill=Batch))+ 
  labs(x=expression(delta^{13}*C[Pro]*"(\u2030)"), y=expression(delta^{13}*C[Hyp]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-30,-10))+
  scale_y_continuous(position = "left",limits=c(-30,-10))+
  geom_point(colour="black", pch=21, size=3.5)+
  #geom_smooth(method=lm, se=FALSE)+
  geom_abline(intercept=c(0), slope= 1, linetype="solid")+
  theme_bw()+
  coord_fixed(ratio = 1)
p
```

#4. Check quality contol through estimated mass balance against observed bulk values.

```{r}
#QC1 d15N mass balance
p <- ggplot(data=data, aes(y=d15N_est, x=d15N, fill=common_name))+ 
  labs(x=expression(delta^{15}*N[OBS]*"(\u2030)"), y=expression(delta^{15}*N[EST]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(5,25))+
  scale_y_continuous(position = "left",limits=c(5,25))+
  geom_point(colour="black", pch=21, size=3.5)+
  geom_abline(intercept=c(0), slope= 1, linetype="solid")+
  theme_bw()+
  coord_fixed(ratio = 1)
p

#QC2 d13C mass balance
p <- ggplot(data=data, aes(y=d13C_est, x=d13C, fill=common_name))+ 
  labs(x=expression(delta^{13}*C[OBS]*"(\u2030)"), y=expression(delta^{13}*C[EST]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-25,-5))+
  scale_y_continuous(position = "left",limits=c(-25,-5))+
  geom_point(colour="black", pch=21, size=3.5)+
  geom_abline(intercept=c(0), slope= 1, linetype="solid")+
  theme_bw()+
  coord_fixed(ratio = 1)
p
```
Compare d13C stdev against error

```{r}
p <- ggplot(data=data, aes(y=nle_d13C_err, x=nle_d13Cstdev, fill= Batch)) + 
  labs(x=expression(delta^{13}*C[stdev]*"(\u2030)"), y=expression(delta^{13}*C[error]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-1,1))+
  scale_y_continuous(position = "left",limits=c(-1,1))+
  geom_point(colour="black", pch=21, size=3.5)+
  theme_bw()+
  guides(color=FALSE, shape=FALSE)+
  coord_fixed(ratio = 1)
p
```



#5. Plot d15N Glu against d15N Phe

```{r}
  p <-ggplot() +
  labs(x=expression(delta^{15}*N[Phe]*"(\u2030)"), y=expression(delta^{15}*N[Glu]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(0,20))+
  scale_y_continuous(position = "left",limits=c(0,40))+
  geom_point(data=human, aes(y=glu_d15N, x=phe_d15N, fill= Culture), colour="black", pch=23, size=3)+
  geom_point(data=fauna, aes(y=glu_d15N, x=phe_d15N, fill = category),  colour ="black", pch=21, size=3.5)+
  theme_bw()+
  annotate("text", label = "T1", x = 7 , y = 0, size = 3, colour = "black",angle = 25)+
  annotate("text", label = "T2", x = 0 , y = 1.8, size = 3, colour = "black",angle = 25)+
  annotate("text", label = "T3", x = 0 , y = 8, size = 3, colour = "black",angle = 25)+
  annotate("text", label = "M2", x = 0 , y = 10.7, size = 3, colour = "Blue",angle = 25)+
  annotate("text", label = "M3", x = 0 , y = 17.6, size = 3, colour = "Blue",angle = 25)+
  annotate("text", label = "M4", x = 0 , y = 24.0, size = 3, colour = "Blue",angle = 25)+
  #scale_fill_manual(values=c(cols), breaks=c(order))+
  theme(legend.title=element_text(size=10))+
  guides(fill=guide_legend(title="Samples"))+
  geom_abline(intercept=c(-8.4,-0.8, 6.8, 9.5, 16.1, 22.7), slope= 1, linetype="dashed")
p
```

Looks ok - plot d15N Glu against d15N Phe. Let's see the relationship between Glu-Phe and d13C. 

```{r}
#Fig 1d Collagen13C against Glu-Phe
p <-ggplot() +
  geom_point(data=human, aes(y=gluphe_d15N, x=d13C, fill = Culture),  colour="black", pch=21, size=3.5)+
  geom_point(data=fauna, aes(y=gluphe_d15N, x=d13C, fill=category), colour="black", pch=23, size=3)+
  labs(y=expression(Delta^{15}*N[Glu-Phe]*"(\u2030)"), x=expression(delta^{13}*C[collagen]*"(\u2030)"))+
  scale_x_continuous(position="top", limits=c(-25,-5))+
  scale_y_continuous(position = "right",limits=c(-5,25))+
  theme_bw()+
   coord_fixed(ratio = 0.5)
p
```





```{r}
#Fig 1c bis Coll 15N against Glu/Phe
p <-ggplot() +
  geom_point(data=human, aes(y=gluphe_d15N, x=d15N, fill=Culture), colour="black", pch=23, size=3)+
  geom_point(data=fauna, aes(y=gluphe_d15N,, x=d15N, fill = category), colour="black", pch=21, size=3.5)+
  labs(x=expression(delta^{15}*N[collagen]*"(\u2030)"), y=expression(Delta^{15}*N[Glu-Phe]*"(\u2030)"))+
  scale_x_continuous(position="top", limits=c(0,20))+
  scale_y_continuous(position = "right",limits=c(-5,25))+
  #scale_fill_manual(values=c(cols), breaks=c(order))+
  theme_bw()
  coord_fixed(ratio = 0.5)
p
```



```{r}
#Val vs Phe d13C
p <- ggplot()+
  labs(x=expression(delta^{13}*C[Val]*"(\u2030)"), y=expression(delta^{13}*C[Phe]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-40,-10))+
  scale_y_continuous(position = "left",limits=c(-40, -10))+
  geom_point(data=human, aes(y=phe_d13C, x=val_d13C, fill = Culture), colour="black", pch=21, size=3.5)+
  geom_point(data=fauna, aes(y=phe_d13C, x=val_d13C,fill=category), colour="black", pch=23, size=3)+
  geom_abline(intercept=c(0), slope= 1, linetype="dashed")+
   #scale_fill_manual(values=c(cols1), breaks=c(order1))+
  theme_bw()+
  coord_fixed(ratio = 1)
p

```




```{r}
p <- ggplot()+
  labs(x=expression(Delta^{13}*C[Lys-Phe]*"(\u2030)"), y=expression(Delta^{13}*C[Val-Phe]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-2,15))+
  scale_y_continuous(position = "left",limits=c(-2, 15))+
  geom_point(data=human, aes(y=valphe_d13C, x=lysphe_d13C, fill = Culture), colour="black", pch=21, size=3.5)+
  geom_point(data=fauna, aes(y=valphe_d13C, x=lysphe_d13C,fill=category), colour="black", pch=23, size=3)+
  geom_abline(intercept=c(0), slope= 1, linetype="dashed")+
   #scale_fill_manual(values=c(cols1), breaks=c(order1))+
  theme_bw()+
  coord_fixed(ratio = 1)+
  theme(legend.title=element_text(size=10))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p
```


```{r}
#Gly vs Phe d13C
p <- ggplot()+
  labs(x=expression(Delta^{13}*C[Val-Phe]*"(\u2030)"), y=expression(Delta^{15}*N[Glx-Phe]*"(\u2030)"))+
  scale_x_continuous(position="bottom", limits=c(-5,10))+
  scale_y_continuous(position = "left",limits=c(-5, 20))+
  geom_point(data=human, aes(y=gluphe_d15N, x=valphe_d13C, fill = Culture), colour="black", pch=21, size=3.5)+
  geom_point(data=fauna, aes(y=gluphe_d15N, x=valphe_d13C, fill=category), colour="black", pch=23, size=3)+
  #geom_abline(intercept=c(0), slope= 1, linetype="dashed")+
   #scale_fill_manual(values=c(cols1), breaks=c(order1))+
  theme_bw()+
  coord_fixed(ratio = 0.5)+
  theme(legend.title=element_text(size=10))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p
```



7# d13C fingerprinting

```{r}
# Install packages
library(googlesheets4)
library(ggplot2)
library (devtools)
library(tidyverse)
library(dplyr)

```

```{r}
# Load colour palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```
#PCA analysis of primary producers

#8. Load the reference data from Larsen

```{r}
gs4_deauth()
#Load Larsen on d13C values for reference
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset primary producters
Larsen_pp <- filter(Larsen, PP == "Y")
#Select groups and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #remove N/As
```
#9. Select just amino acids for PCA
```{r}

all_AA <- Larsen_pp %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Plot PC1 vs PC2
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") +
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  coord_fixed (ratio = 1)
p
```
10. Normalise the AA d13C values 
```{r}
#Calculate the AAn values which equals each values variance from the mean and repeat the analysis.

#First organise the data and normalise to row mean.
Larsen_pp <- filter(Larsen, PP == "Y")
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. .
Larsen_pp <- Larsen_pp %>% rowwise() %>% mutate(row_mean = mean(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% rowwise() %>%
  mutate(across(c(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C), ~ .x - row_mean))

```
#10. Then perform the PCA and visualise the results. This creates Figure 1 in Larsen et al.
```{r}

#extract the amino acid values
all_AA <- Larsen_pp_n %>%  dplyr::select(ala_d13C, asx_d13C, glx_d13C, gly_d13C, ile_d13C, thr_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C, tyr_d13C)
PCA <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues <- data.frame(group = Larsen_pp_n$Group, PCA$x)
# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(PCA$rotation), PCA$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") +
  annotate ("text", x = (PCAloadings$PC1*11), y = (PCAloadings$PC2*11),
     label = PCAloadings$Variables, )+
xlim(5,-5)+
 ylim(5,-5)+
  coord_fixed (ratio = 1)
p
```
#11. Repeat PCA just for source aminios
```{r}

Larsen_ppS <- filter(Larsen, PP == "Y")
Larsen_ppS <- Larsen_pp %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_ppS <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. .
Larsen_ppS <- Larsen_ppS %>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
Larsen_pp_Sn <-Larsen_pp %>% rowwise() %>%
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
#Conduct the PCA
all_AA <- Larsen_pp_Sn %>%  dplyr::select(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
PCA1 <- prcomp(all_AA, scale. = TRUE)
# Extract PC axes for plotting
PCAvalues1 <- data.frame(group = Larsen_pp_Sn$Group, PCA1$x)
# Extract loadings of the variables
PCAloadings1 <- data.frame(Variables = rownames(PCA1$rotation), PCA1$rotation)
# Visuqlise the PCA
p <- ggplot (data=PCAvalues1, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size=3) +
  geom_segment(data = PCAloadings1, aes(x = 0, y = 0, xend = (PC1*10),
     yend = (PC2*10)), arrow = arrow(length = unit(1/2, "picas")),
     color = "black") +
  annotate ("text", x = (PCAloadings1$PC1*11), y = (PCAloadings1$PC2*11),
     label = PCAloadings1$Variables, )+
xlim(10,-10)+
 ylim(10,-10)+
  coord_fixed (ratio = 1)
p
```
#12. LDA analaysis

Prepare Larsen data for LDA analysis
```{r}
#Select just amino acids for PCA
library(caret)
library(tidyverse)
#Create a normalised training set from Larsen's primary producters
gs4_deauth()
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset non primary producers
Larsen_pp <- filter(Larsen, PP == "Y")
#Select group and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_pp <- na.omit(Larsen_pp)
Larsen_pp$Group  <-as.factor (Larsen_pp$Group)
#omit Fungi
#Larsen_pp <-Larsen_pp %>% filter(Group!="Fungi")

```


Normalise the amino acids in Larsen reference dataset 
```{r}
#Create a new coloumn containing the row mean of the key amino acid for each row. .
Larsen_pp <- Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(across(c( ile_d13C, leu_d13C,  lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
Larsen_pp_n <- Larsen_pp_n %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)#drop means 
print (Larsen_pp_n)

```

Partition the data and create train and test sets 

```{r}
ind <- sample(2, nrow(Larsen_pp_n),
              replace = TRUE,
              prob = c(0.9, 0.1))
train.data <- Larsen_pp_n[ind==1,]
test.data <- Larsen_pp_n[ind==2,]
```

Create LDA model
```{r}
library(MASS)
# Fit LDA model to the training data
lda_model <- lda(Group ~ ., data = train.data)
```
Evaluate the LDA  performance
```{r}
# Predict on the test data using the LDA model
lda_predictions <- predict(lda_model, test.data)

#lda_predictions$class
confusion_matrix <- confusionMatrix(lda_predictions$class, test.data$Group)
print(confusion_matrix)
```

Call arcaheological data and normalise
```{r}
data <- all_data %>%  dplyr::select(category, Culture, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
data <- data %>% dplyr::rowwise() %>% dplyr::mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))

#Subtract the row mean from each of the row values to normalize
data <- data %>% dplyr::rowwise() %>% dplyr::mutate(across(c( ile_d13C, leu_d13C,  lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
data_n <- data %>%  dplyr::select(category, Culture, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)#drop means 
data_n <- data_n %>% rename(Group=category) 
data_n$Group <-as.factor (data_n$Group)
print (data_n)
```

```{r}
# Predict the classes for the unknown samples using the LDA model
lda_Danish_predictions <- predict(lda_model, newdata = data_n)
#Print the predicted classes for the unknown samples
predicted_Danish_classes <- lda_Danish_predictions$class
#print (predicted_Danish_classes)
```
Create two dataframes for plotting and combine

```{r}
Refs <- data.frame(LD1 = predict(lda_model, newdata = train.data)$x[, 1], LD2 = predict(lda_model, data = train.data)$x[, 2], Type = "Training", group = as.factor(train.data$Group))
Samples <- data.frame (LD1 = lda_Danish_predictions$x[, 1], LD2 = lda_Danish_predictions$x[, 2], Type = "Samples", group = data_n$Group, culutre = data_n$Culture)
#Combined <- rbind (Refs, Samples)
#print (Combined)
```



#Find means for automatic labelling
``` {r}
#Find means for automatic labelling
means <- Refs %>% group_by(group) %>%
  summarise(dplyr::across(c(LD1, LD2), mean))
means
#means <- means %>%mutate(cats=c("Porcine", "Ruminant", "Dairy", "Freshwater", "Marine"))#Sh
```


Plot the LDA
```{r}
# Plot the combined data
p <- ggplot() +
 #geom_point(data=Refs, aes(x =LD1, y =LD2, fill = group), pch=22, colour = "black", size = 2) +
  stat_ellipse(geom= "polygon", data=Refs, alpha=0.1,
               aes (x = LD1, y = LD2, group = group),
              level = 0.68)+
  geom_point(data=Samples, aes(x =LD1, y =LD2, fill = group), pch=21, colour = "black", size = 3) +
  geom_text(data=means, aes(y=LD2, x=LD1,label=group), hjust=0, size=4, color="black")+
  labs(title = "LDA Plot", x = "Linear Discriminant 1", y = "Linear Discriminant 2", fill = "Group", shape = "Type") +
  theme_bw()
p
```














```{r}
library(tidyverse)
library(caret)
```
#Create a normalised test data set from Larsen

```{r}

```
Creat a normalised training set from Larsen's primary producters
```{r}
Larsen <<- read_sheet(ss = "https://docs.google.com/spreadsheets/d/19YG0RAg8fuNAKWujooXnLnv6iNRheXXoKT331Klac-I/edit#gid=1182477512", sheet="Larsen")
#Subset non primary producers
Larsen_pp <- filter(Larsen, PP == "Y")
#Select groups and amino acids
Larsen_pp <- Larsen_pp %>%  dplyr::select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
Larsen_pp <- na.omit(Larsen_pp) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
Larsen_pp <- Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))
#Subtract the row mean from each of the row values to normalize
Larsen_pp_n <-Larsen_pp %>% dplyr::rowwise() %>% dplyr::mutate(across(c( ile_d13C, leu_d13C,  lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
Larsen_pp_n
```
Creat a test set from TRADITION data

```{r}
data <- data %>%  select(Group, ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
data <- na.omit(data) #
#Create a new coloumn containing the row mean of the key amino acid for each row. . 
data <- data%>% rowwise() %>% mutate(row_mean = mean(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)))#
#Subtract the row mean from each of the row values to normalize
data_n <-data %>% rowwise() %>% 
  mutate(across(c(ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C), ~ .x - row_mean))
data_n
```




```{r}

# Create 2 dataframes for comparison
train.data <-  Larsen_pp_n %>%  dplyr::select(Group,  ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)

test.data <- data_n %>%  dplyr::select(Group,  ile_d13C, leu_d13C, lys_d13C, val_d13C, phe_d13C)
```

```{r}

# Estimate preprocessing parameters
preproc.param <- train.data %>% 
  preProcess(method = c("center", "scale"))
# Transform the data using the estimated parameters
train.transformed <- preproc.param %>% predict(train.data)
test.transformed <- preproc.param %>% predict(test.data)
test.transformed
```

LDA determines group means and computes, for each individual, the probability of belonging to the different groups. The individual is then affected to the group with the highest probability score.

```{r}
library(MASS)
# Fit the model
model <- lda(Group~., data = train.transformed)
# Make predictions
predictions <- predict(model,test.transformed)
# Model accuracy
mean(predictions$class==test.transformed$Group)
model <- lda(Group~., data = train.transformed)
model
plot(model)
```

Make predictions

```{r}
predictions <- model %>% predict(test.transformed)
names(predictions)
# Predicted classes
head(predictions$class, 6)
# Predicted probabilities of class memebership.
head(predictions$posterior, 6) 
# Linear discriminants
head(predictions$x, 3) 
```
Plot results
```{r}
lda.data <- cbind(train.transformed, predict(model)$x)
ggplot(lda.data, aes(LD1, LD2)) +
  geom_point(aes(color = Group))
mean(predictions$class==test.transformed$Group)
lda.data
```

